---
title: "First steps with dada2pp"
author: "Pedro Martinez Arbizu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First steps with dada2pp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Working with dada2pp
by *Pedro Martinez Arbizu*


## Starting point
You have analysed your metabarcoding results with dada2 (or other pipeline) and produced a
 *community table*. This means you have ASVs (or OTUs) as lines and localities as columns.

## Getting ready
install library from Github

*Windows users need Rtools installed to use devtools*
```{r,eval = FALSE}
library(devtools)
install_github("pmartinezarbizu/dada2pp")
```


load library
```{r}
library(dada2pp)
```

## Create a color palette
The function `pal2table()` created a matching table of taxa to colors and symbols, so that you can
assign same colors and symbols to your taxa consistently in all your graphs. 
```{r}
data(iris)
library(dada2pp)
iriscol <- pal2table(unique(iris$Species))
```
inspect the created object:
```{r}
iriscol
```
View the assigend colors and sysmbols:
```{r, fig.show='hold'}
plot(iriscol)
```



- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

```{r, echo=FALSE, results='asis'}
knitr::kable(iriscol)
```

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
