---
title: "First steps with dada2pp"
author: "Pedro Martinez Arbizu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First steps with dada2pp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Working with dada2pp
by *Pedro Martinez Arbizu*

> "This tutorial should be present
<a href="https://github.com/pmartinezarbizu/dada2pp/wiki" target="_blank">here</a>"



## Starting point
You have analysed your metabarcoding results with dada2 (or other pipeline) and produced a
 *community table*. This means you have ASVs (or OTUs) as lines and localities as columns.

## Getting ready
install library from Github

*Windows users need Rtools installed to use devtools*
```{r,eval = FALSE}
library(devtools)
install_github("pmartinezarbizu/dada2pp")
```


load library
```{r}
library(dada2pp)
```

Load the example dataset `deepMeio`.
This dataset contains the result of metabarcoding of deep-sea meiofauna.
The amplified fragment was V1V2 from 18s. The dada2 AVS were blasted against GenBank and
 some columns with statistic results were added (columns 1:8).
 The number of reads on 19 stations is recorded on colums 9:27. Each line represents one ASV.

```{r}
data(deepMeio)
``` 

Inspect the dataset. 
```{r}
summary(deepMeio)
```

Before you move on, it would be useful to declare some information on `gene` (fragment studied),
`pip` (pipeline used) and `pr` (project name). This will help to consistent names for files and graphs.

```{r}
gene <- 'V1V2'
pip <- 'dada2'
pr <- 'DeepMeio' 
```
## Filter some ASVs
Produce a graph of frament lengths

```{r, fig.show='hold'}
mode <- as.integer(names(which(table(deepMeio$length)==max(table(deepMeio$length)))))
counts <- max(table(deepMeio$length))
plot(table(deepMeio$length),ylab='number of  ASVs',xlab='read length')
text(mode,counts,labels=paste (mode, 'bp'),pos = 4)
```

Apply filters if needed:
- 1.) remove ASVs shorter than (for instance) 300bp
```{r}
deepMeio <- deepMeio[deepMeio$length>=300,]
```
Other filter could be to remove ASV with query coverage `qcovs` is lower than (for instance) 90%.
Try this:
```{r,eval=FALSE}
plot(table(deepMeio$qcovs))
deepMeio <- deepMeio[deepMeio$qcovs>=90,]
```	
	

- 2.) remove samples with very low number of reads if needed.

	First inspect number of reads per sample:

```{r}
apply(deepMeio[,9:27],FUN=sum,MARGIN=2)
```
As there is no sample with extrem low number of reads we keep all.

Otherwise, to remove (for instance) sample loc1_2:
```{r, eval = FALSE}
deepMeio <- deepMeio[,-which(names(deepMeio) == 'loc1_2')]
```

After applying all filters, check if any higher taxon group was left with 0 ASVs:
```{r}
table(deepMeio$Group)
```
This is not the case here. Otherwise remove the groups with 0 ASVs.
To remove for instance group `mesozoans` (if it would have 0 ASVs):

```{r,eval=FALSE}
deepMeio <- deepMeio[-which(deepMeio$Group == 'mesozoans'),]
```

And then drop the level `mesozoans` from `Group`:

```{r}
deepMeio$Group <- droplevels(deepMeio$Group)
```
 


## Clean up Genbank inconsistent classification

Column `Group` contains the higher taxon assigment for the AVS. How many are different groups are there?
```{r}
unique(deepMeio$Group)
```
At first glace we discover 3 inconsistecies:

- There is a group `crustaceans` but also a separate group `isopods` which are obviously also crustaceans.
- There is a separate group `animals`
- There is a separate group `eukaryotes`

Assign the isopods to the group `crustaceans`:
```{r}
deepMeio[deepMeio$Group == 'isopods', 'Group'] <- 'crustaceans'
```
Inspect which taxa are in the group `animals`:
```{r}
unique(deepMeio[deepMeio$Group == 'animals', 'Species'])
```
Surprisingly *Haplogonaria*, *Ascoparia*, *Nemertoderma* and *Proporus* are all `flatworms`.

Move these ASVs to `flatworms`:
```{r}
deepMeio[deepMeio$Species %in% c( "Haplogonaria_sp._'schillingi'_UJ2011",
 "Ascoparia_sp._UJ2011", "Nemertoderma_sp._SMNHUJ13330", "Proporus_brochii"), 'Group'] <- 'flatworms'
```
check results with these two commands:
```{r, eval = FALSE}
unique(deepMeio[deepMeio$Group == 'animals', 'Species'])
unique(deepMeio[deepMeio$Group == 'flatworms', 'Species'])
```

Inspect which taxa are in the group `eukaryotes`:
```{r}
unique(deepMeio[deepMeio$Group == 'eukaryotes', 'Species'])
```
As none of the taxa is metazoan meiofauna we leave the group `eukaryotes` as it is.


## Define your target group
Our study focus on free-living metazoan meiofauna. From the remaining groups there are some of them
 which could be defined as contamination, others are benthic deep-sea organisms but not our
 target metazoan meiofauna and only some groups are what we are looking for.
 We define 3 groups `contamination`, `non_target` and `target`:

```{r}
contamination <- c('animals', 'eukaryotes', 'dinoflagellates', 'ascomycetes','golden_algae') 

non_target <- c('cercozoans', 'hemichordates', 'hydrozoans', 'bryozoans', 'protozoa', 'mesozoans',
'sea_cucumbers', 'tunicates', 'tusk_shells', 'solenogasters')

target <- c('segmented_worms', 'crustaceans', 'ribbon_worms', 'nematodes', 'bivalves',
'flatworms', 'gastrotrichs', 'gastropods', 'loriciferans', 'spoonworms')
```

Use the function `stats.tt` to inspect your results and reduce your dataset to target taxa.
- x: your data.frame
- reads: the subset of columns containing the reads
- by: the column containig the grouping factor
- cont: the contamination levels
- non_target: the non target levels

Everything that is not contamination or non_target is assumend to be target:

```{r}
stat <- stats.tt(x=deepMeio,reads=deepMeio[,9:27],by=deepMeio$Group,cont=contamination,non_target=non_target)

stat$stats
```
We can see form the output of `stat$stats` that the target group has 10 taxa and retains
 most of the reads (4.4 Million) and ASVs (n_otus = 836).

Inspect the stat object: 
```{r, eval = FALSE}
stat$cont
stat$non_target
stat$target
```

Write stats table to a file:
```{r, eval = FALSE}
write.table(stat$stats,file=paste('target.stats',gene,pip,pr,'.csv',sep='_'),sep=';',row.names=TRUE,col.names=NA)
```
Extract your target ASVs to new dataset:
```{r}
meio <- stat$target.table
```
remove non used taxa from `Group`
```{r}
meio$Group <- droplevels(meio$Group)
```



## Create a color palette
Before we proceed with producing graphs it would be good to have consistent colors for the taxon groups in all the graphs.
The function `pal2table()` creates a matching table of taxa to colors and symbols, so that you can
assign same colors and symbols to your taxa consistently in all your graphs. 
Check `?pal2table()` for more options and control.
```{r}
meiocol <- pal2table(unique(meio$Group),pal='c25')
```
inspect the created object:
```{r}
meiocol
```
View the assigend colors and sysmbols:
```{r, fig.show='hold'}
plot(meiocol)
```

## Use your color palette
To use your color palette there is the function `match2table`. See `?match2table()` for details.
```{r, fig.show='hold'}
t1 <- table(meio$Group)
or <- order(t1,decreasing=TRUE)
assign.col <- match2table(names(t1[or]),meiocol,'col')
par(oma = c(4, 0, 0, 0),xpd=NA)
barplot(t1[or], col=assign.col,las=2)
```

## Create graph with number of reads and ASV per sample
The function `barp.table` will create the information needed to produce barplots.
Inspect the results in `..$table`. See `?barp.table()` for more details.
```{r}
bp <- barp.table(meio[,9:27],meio$Group,sum) 

bp$table
```

Create the barplots (see `?plot.barp_table()` for details):
```{r,fig.width = 13,fig.height = 10}
plot(bp,'dada2sum',pal2table=meiocol,ncol.leg=5,oma1=7)
```



## Statistics for comparing between locations
In the following section we will explain how to compare between locations from which we have replicate samples.
We have 7 samples in location 1 and 12 samples in location 2. Create a vector:
```{r}
locations <- c(rep('loc1',7),rep('loc2',12))
```
Create a color palette for the locations:
```{r,fig.width = 5,fig.height = 4}
loccol <- pal2table(unique(locations))
plot(loccol)
```
 
## Count shared ASVs between locations
Use the function `countTaxa` to performe the statistics
- x: the community table
- taxa: the vector containing the taxa to count
- fac: the factor grouping the samples

```{r}
countTaxa(x=meio[,9:27],taxa=meio$Group, fac=locations)
```

## Diversity values between locations
Function 

```{r}
dd <- diverse.tt(x=meio[,9:27],by=locations,pal=loccol)

dd
```

```{r}
plot(dd)
```
